---
title: "Class06:Homework Function"
author: "Bianca Barriga"
format: pdf
---
The provided code that we want to improve and adapt to be more general for 
# Can you improve this analysis code? library

```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug 
s2 <- read.pdb("1AKE") # kinase no drug 
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA") 
s2.chainA <- trim.pdb(s2, chain="A", elety="CA") 
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")

s1.b <- s1.chainA$atom$b 
s2.b <- s2.chainA$atom$b 
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor") 
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```
Plot 1 and 3 are the same. correct the code 

```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug 
s2 <- read.pdb("1AKE") # kinase no drug 
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA") 
s2.chainA <- trim.pdb(s2, chain="A", elety="CA") 
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")

s1.b <- s1.chainA$atom$b 
s2.b <- s2.chainA$atom$b 
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor") 
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

Now simplify the code - delete the lines that look the same, the only difference was the input 
```{r}
library(bio3d)

id <- "4AKE"
s1 <- read.pdb(id) # kinase with drug 

s1.chainA <- trim.pdb(s1, chain="A", elety="CA") 

s1.b <- s1.chainA$atom$b 

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")

```

#code extract function 
```{r}
library(bio3d)

#' @param_id The PDB code of your unput structure
#' 
#' @return A plot of B factor values
#' 
#' @ export
#' 
#' @examples pdblot("AKE")

id <- "1AKE"
pdbPlot <- function(id) {
  #read the input PDB file
  s1 <- read.pdb(id) # kinase with drug 
  #Extract a subset of chain A
  chain <- trim.pdb(s1, chain="A", elety="CA") 
  #
  #s1.b <- s1.chainA$atom$b 
  #plot
  plotb3(chainA$atom$b, sse=chain, typ="l", ylab="Bfactor")
}

```

#use apply function over columns and rows where Ids are listed to apply the function over the multiple ids in for the dataset. 

pdb <- function(id="1E4Y") {
s1<- read.pdb(id)
chain <- trim.pdb(s1, chain="A", elety ="CA")
plotb3(chainA$atom$b, sse=chain, typ="l", ylab="Bfactor")
}

>Q1. What type of object is returned from the read.pdb() function?


